---
title: MADA Fall 2019 - Topic Introduction 
subtitle: ""
author: Andreas Handel
institute: "University of Georgia"
date: "`r Sys.Date()`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
---

```{r, echo = FALSE}
#just so I can use emoticons
#devtools::install_github("hadley/emo")
library(emo)
```

# Introduction
Since you are taking this class, you likely know what data analysis is all about. I therefore do not need to describe it much - we'll get into the details soon anyway. Instead, I'm providing a few hopefully inspiring and motivating examples of data analysis and use. You'll be doing similar projects yourself soon!


# Motivating Videos

There are lots of good videos online discussing different aspects of the vast topic of data analysis. TED is an especially good source for such videos.

## TED talk 1
Arguably one of the best communicators of Public Health data was the late [Hans Rosling](https://en.wikipedia.org/wiki/Hans_Rosling). He has several great entertaining and enlightening TED talks. Here is one of them - a statistics talk with >13 million views!

<p>
<iframe src="https://embed.ted.com/talks/hans_rosling_at_state" width="640" height="360" frameborder="0" scrolling="no" webkitAllowFullScreen mozallowfullscreen allowFullScreen></iframe>
</p>

You can watch [more TED talks by Hans Rosling here](https://www.ted.com/speakers/hans_rosling). The software he uses to illustrate data is available online on the [Gapminder website](https://www.gapminder.org/).


## TED talk 2
In this entertaining and empowering talk, [Talithia Williams](https://www.ted.com/speakers/talithia_williams) argues for the utility of careful assessment of data in your personal life.

<p>
<iframe width="560" height="315" src="https://www.youtube.com/embed/_GMVTJ9ZKVc" frameborder="0" allowfullscreen></iframe>
</p>


## TED talk 3
In this talk, [Sebastian Wernicke](https://www.ted.com/speakers/sebastian_wernicke) gives some cautionary suggestions about trusting data too much and argues for critical assessment as an important part of any data analysis and decision process.

<p>
<iframe src="https://embed.ted.com/talks/sebastian_wernicke_how_to_use_data_to_make_a_hit_tv_show" width="640" height="360" frameborder="0" scrolling="no" webkitAllowFullScreen mozallowfullscreen allowFullScreen></iframe>
</p>


## More TED talks 
The TED website has a large number of data related talks, [which you can check out here](https://www.ted.com/topics/data). Feel free to explore as much as you like. 


# Case studies
It is by now very common that individuals post data analyses online in a reproducible manner, i.e. providing all the code and data to recreate it.

## TidyTuesday
An interesting initiative is [TidyTuesdays](https://thomasmock.netlify.com/post/tidytuesday-a-weekly-social-data-project-in-r/), where every Tuesday [a new dataset is released here](https://github.com/rfordatascience/tidytuesday) and individuals are encouraged to analyze it. Later in this course, you will be participating in some TidyTuesdays yourself.

Individuals who participate in TidyTuesday often post their projects online and link to it on social media. See for instance the Twitter posts at the bottom of the [TidyTuesdays announcement site](https://thomasmock.netlify.com/post/tidytuesday-a-weekly-social-data-project-in-r/).

Here is a nice example of [David Robinson](http://varianceexplained.org/) analyzing a wine rating dataset. He does a live recording of him digging into the data. 

<p>
<iframe width="560" height="315" src="https://www.youtube.com/embed/AQzZNIyjyWM" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</p>

You don't need to watch the whole 85 minutes, but do watch at least the first 15 minutes to see him coding/analyzing in real time. You might just get hooked and keep watching `r emo::ji("smile")`. It should be interesting even if you don't care about wine. If you have no R experience, you will not understand exactly the commands he types in. You will learn those commands very soon. You will still get the overall idea of how he does the analysis. And his thinking out loud is informative as well. 

Dave also does his work reproducibly and shares all the code. See the links provided in the video description. For convenience, [here is the link to download the code for his analysis](https://raw.githubusercontent.com/dgrtwo/data-screencasts/master/wine-ratings.Rmd). Just save the file after it opened in your browser. It should be named `wine-ratings.Rmd`. Click on it, it should open in RStudio (if you don't have R/RStudio installed yet, go through module 2, then come back). 

Once the file is open in RStudio, you might see at the top of the file a message suggesting that you need to install several packages to run the file. Do so. If you don't get that auto-suggestion, you need to install the pacakges by hand. Every package that is called with the `library()` commmand needs to be installed. There is one package that is only available for Linux (and maybe Mac?) computers. So we need to alter the code a bit to make things work for everyone. Find these lines of code:

```{r, eval = FALSE}
library(doMC)
registerDoMC(cores = 4)
````

and add a `#` in front of them to turn them into comments and therefore disable them (you could also delete these lines, but then they are gone for good, so disabling with `#` is often better).

Once you have done that and installed all other packages, hit the `Knit` button. If you are still missing a package, you get an error message. If things work out, the whole script will run a while (depending on the speed of your computer), and you should get a document/report which reproduces the complete analysis he did in the video! 

Again, most or all of the R code might not make sense to you just yet. That's ok. The main point is to show you a nice example of a data analysis that is done in a way that others (you) can reproduce it with just a few clicks.

## XX



