---
title: "Modern Applied Data Analysis"
output: 
  html_document:
    toc: FALSE
---

```{r, include = FALSE}
#just so I can use emoticons
#devtools::install_github("hadley/emo")
library(emo)
```


Welcome to the **Modern Applied Data Analysis (MADA)** course website.

This website contains almost all the content of an online course taught in Fall 2019 at the University of Georgia by [Andreas Handel](https://www.andreashandel.com/). Most materials are hosted here, some are on external sites and either linked or embedded from here.

I plan to keep this website continuously online, with updates happening every time I teach the course again. 

# Website content

The __General Information__ section provides a class __Syllabus__ and a __Schedule__. The latter will likey be updated throughout the course, so check it regularly.

The majority of the content is listed under the __Modules__ heading. Each module has several units in a drop-down/pop-out menu. Go through them in the listed order.

Instructions for specific exercises are listed in the __Exercises__ section. You will be told when you should access those documents and do the exercises. Of course, reading ahead is allowed `r emo::ji('smiley')`.

As part of the class, you will be doing a semester-long data analysis project. Instructions and detailed information for this are given in the __Project__ section.

The __Communication__ section provides information on how course communications are planned to happen.

The __Glossary__ section provides brief definitions of terms used in this course. 

The __Resources__ section contains a listing of external materials. Some of this material will assigned as part of this course. You will be told what exactly to read or work through in the modules documents. Other resources are listed in case you might find them helpful. All materials are (should be) available freely online.

Note that by default, all links open in the current tab. You'll likely want external ones to open in a separate tab or window. To do so, you can (with most browsers) hold the `Ctrl` button when clicking, or do a right-click and select `open in new tab/window`.


# Getting started

I suggest you start by looking over the __Syllabus__ and __Schedule__ documents, then __Communication__ and __Resources__. After that, start with the __Module 1__ content in the given order. You can read through the _Project_ section early or wait until you have gone through at least modules 1 and 2.

## Acknowledgements

This course was developed and is maintained by [Andreas Handel](https://www.andreashandel.com/).

```{r, results = "asis", echo = FALSE, message = FALSE, warning=FALSE}
#the following code snipped is taken from Hadley's R4DS intro page
library(dplyr)
contributors <- readr::read_csv("contributors.csv", col_types = list())
contributors <- contributors %>% 
  mutate(
    link = glue::glue("[\\@{GithubID}](https://github.com/{GithubID})"),
    desc = ifelse(is.na(LastName), link, glue::glue("{FirstName} {LastName} ({link})"))
  )
cat("The following individuals have contributed to improving the course:\n")
cat(paste0(contributors$desc, collapse = ", "))
```

The course materials are licensed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-nc-sa/4.0/). Linked and embedded materials are governed by their own licenses. I assume that all external materials used or embedded here are covered under the educational fair use policy. If this is not the case and any material displayed here violates copyright, please let me know and I will remove it.


  
