---
title: "Modern Applied Data Analysis"
output: 
  html_document:
    toc: FALSE
---


```{r, include = FALSE}
#just so I can use emoticons
#only load here, not on each page separately
library(emoji)
```


<!-- Some trickery to get the logo to show -->
<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"media/MADAlogo.png\" style=\"float: right; border: none; width: 150px;\"/>')
   });
</script>


Last updated: `r Sys.time()`


<!-- _**The website will be continuously updated throughout the semester with new material released on an ongoing basis, as well as adjustments/fixes as needed. Visit/check regularly, especially the Schedule. Note the last update time above.**_ -->


Welcome to the **Modern Applied Data Analysis (MADA)** course website.

This website contains almost all the content of an online graduate level course taught by [Andreas Handel](https://www.andreashandel.com/) at the University of Georgia. Most materials are hosted here, some are on external sites and either linked or embedded from here.

When the course is running, some material might not be visible and will be unlocked as the course proceeds. The course ran last in fall 2021. At this point, all materials are online and will remain so with likely only small updates happening on an ongoing basis and larger updates expected once the course runs again. 


# Website content

The __General Information__ section provides a class __Syllabus__ and a __Schedule__. The latter will likely be updated throughout the course, so check it regularly. The __Communication__ page provides information on how course communications are planned to happen. The __Assessments__ explains how the different assignments and assessments for the course work.

The majority of the content is listed under the __Modules__ heading. Each module has several units in a drop-down/pop-out menu. Go through them based on the instructions in the __Schedule.__ Each module ends with a document listing the assessments that go with the module.

The __Project__ section provides instructions and detailed information for the semester-long data analysis project which is part of the course. 

The __Resources__ section contains a __Course Resources__ section with listings of materials that are most relevant to the course. Some of this material will assigned as part of this course. You will be told what exactly to read or work through in the modules documents. The __General Resources__ section lists other resources in case you might find them helpful. All materials are (should be) available freely online. The __Glossary__ section provides brief definitions of terms used in this course. 

Note that by default, all links open in the current tab. You'll likely want external ones to open in a separate tab or window. To do so, you can (with most browsers) hold the `Ctrl` button when clicking, or do a right-click and select `open in new tab/window`.


# Getting started

Start by looking over the documents in the __General Information__ section, then the __Resources__ section. After that, start with the first __Module__ content in the given order. You can read through the __Project__ section early or wait until you have gone through the first few modules.


## Acknowledgements

This course was developed and is maintained by [Andreas Handel](https://www.andreashandel.com/).

```{r, results = "asis", echo = FALSE, message = FALSE, warning=FALSE}
#the following code snipped is taken from Hadley's R4DS intro page
library(dplyr)
contributors <- readr::read_csv("contributors.csv", col_types = list())
contributors <- contributors %>% 
  mutate(
    link = glue::glue("[\\@{GithubID}](https://github.com/{GithubID})"),
    desc = ifelse(is.na(LastName), link, glue::glue("{FirstName} {LastName} ({link})"))
  )
cat("The following individuals have contributed to improving the course:\n")
cat(paste0(contributors$desc, collapse = ", "))
```

<!-- Some icons come from [Desirée De Leon](https://desiree.rbind.io/post/2019/making-tip-boxes-with-bookdown-and-rmarkdown/). -->

The course materials are licensed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-nc-sa/4.0/). Linked and embedded materials are governed by their own licenses. I assume that all external materials used or embedded here are covered under the educational fair use policy. If this is not the case and any material displayed here violates copyright, please let me know and I will remove it.


  
