---
title: A testbed for Shiny
subtitle: ""
author: Andreas Handel
institute: "University of Georgia"
date: "`r as.Date(file.mtime(knitr::current_input()))`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc: FALSE
---



```{r, echo = FALSE}
#just so I can use emoticons
#devtools::install_github("hadley/emo")
library(emo)
```

This page currently only exists so I can test embedding of Shiny apps and other interactive content.

# DSAIDE

This is an embedding of the DSAIDE R package. You should not use it through this website. I suggest you either go directly to its online location [here](https://shiny.ovpr.uga.edu/DSAIDE/) or install it locally. [See more here.](https://ahgroup.github.io/DSAIDE/)

```{r echo=FALSE,  scrolling="no", width="100%"}
knitr::include_url("https://shiny.ovpr.uga.edu/DSAIDE/")
```


# LearnR

This embeds a learnR bit of code. Content not related to this course, just a test.

```{r echo=FALSE,  scrolling="no", width="100%"}
set.seed(12)
heights_island1 <- rnorm(50,10,2)
knitr::include_url("https://tinystats.shinyapps.io/06-standardError-ex5/")
```

