---
title: "Coding Exercise Solution"
output:
  html_document:
    df_print: paged
---


Load packages. You need to install them if they are not present.

```{r}
library(dslabs)
```


We'll use the gapminder data. We'll first take a look at it. The help for the data will not show up in the notebook but should be coming up in the bottom right RStudio window.
```{r}
help("gapminder")
str(gapminder)
summary(gapminder)
class(gapminder)
```

# Base R practice

The following uses base R commands. 

We extract the African countries and save them as a new object. We also look at them using `str` and `summary`.
```{r}
africadata <- gapminder[gapminder$continent == 'Africa',]
str(africadata)
summary(africadata)
```

Now we make 2 variables/data frames that contain only infant_mortality+life_expectancy and infant_mortality+population.
```{r}
dat1 <- africadata[,c("infant_mortality","life_expectancy")]
dat2 <- africadata[,c("population","life_expectancy")]
str(dat1)
summary(dat1)
str(dat2)
summary(dat2)
```

Using base plot commands, we create scatterplots for each dataset, using a log scale for the x-axis in the second plot.
```{r}
plot(dat1$infant_mortality,dat1$life_expectancy,type = 'p')
plot(dat2$population,dat2$life_expectancy,type = 'p',log = 'x')
```

Find years for which infant mortality is missing
```{r}
missingyears <- africadata$year[is.na(africadata$infant_mortality)]
print(missingyears)
```

Extract data for year 2000
```{r}
africa2000 <- africadata[africadata$year == 2000,]
str(africa2000)
summary(africa2000)
```

Plot data for year 2000
```{r}
plot(africa2000$infant_mortality,africa2000$life_expectancy,type = 'p')
plot(africa2000$population,africa2000$life_expectancy,type = 'p',log = 'x')
```

Fit linear model using life expectancy as outcome and mortality or population as predictors. We can look at the results from the fit using the `summary` function.
```{r}
fit1 <- lm(africa2000$life_expectancy ~ africa2000$infant_mortality)
fit2 <- lm(africa2000$life_expectancy ~ africa2000$population)
summary(fit1)
summary(fit2)
```

