---
title: "Coding Exercise Solution"
output:
  html_document:
    df_print: paged
---



# Tidyverse practice

We'll what we did above, but now we are using functionality from some of the packages in the `tidyverse` and others.

We'll load the `tidyverse` collection of packages. If you don't have them installed, do so. This will likely install a lot of packages and take a while. We'll also load the `skimr` package. Install as needed.

```{r}
library(dslabs)
library(tidyverse)
library(skimr)
```

_Note that while I'm only loading these packages here for didactic reasons, it is generally best to have all `library` commands at the beginning of a document, this way a user can quickly see which ones to install. That prevents someone running a long simulation, only to get an error message at the end because a package to produce the final plot is missing._ 

```{r}
dplyr::glimpse(gapminder)
skimr::skim(gapminder)
```

Pull out african countries and only some of the variables. 

```{r}
africasmall <- gapminder %>% dplyr::filter(continent == "Africa") %>% dplyr::select(life_expectancy,country,population,infant_mortality) 
skim(africasmall)
```


Note that I often write both the function name and the package where the function comes from, i.e. `dplyr::filter` instead of just `filter`. It is not strictly needed, but if multiple functions with the same name exist in different packages (which is for instance true for `filter`), being explicit about which one you want helps with potential problems in the code.


Plotting life expectancy as function of infant mortality and population, giving each country a separate color.

```{r}
p1 <- africasmall %>% ggplot(aes(x = infant_mortality, y = life_expectancy, color = country)) + geom_point()
plot(p1)
p2 <- africasmall %>% ggplot(aes(x = population, y = life_expectancy, color = country )) + geom_point() + scale_x_log10()
plot(p2)
```
```{r}
gapminder %>% dplyr::filter(continent == "Africa", year == 2000) %>%  
             ggplot(aes(x = infant_mortality, y = life_expectancy)) + geom_point() + geom_smooth(method = "lm") -> p3
plot(p3)
```

